NumLin: Linear Types for Linear Algebra

### 2019-01-11 write-up/paper/discussion.tex
Typo fix

Using linear types for BLAS routines is a particularly good domain fit
(given the implicit restrictions on aliasing arguments), and as a
result the idea of using substructural types to express array
computations is not a particularly new
one\cite{scholz,henriksen,bernardy2016}.  However, many of these
designs have been focused on building languages to \emph{implement}
the kernel linear algebra functions, and as a result, they tend to add
additional limitations on the language design. Both
Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
higher-order functions facilitating compilation to GPUs. The work of
\cite{bernardy2016} forbids term-level recursion, in order to ensure
that all higher-order computations can be statically normalized away
and thereby maximize opportunities for array fusion.

@@

Using fractional permissions directly, as we did, rather than relying
on a complex has two main benefits. First, our type system
demonstrates that type systems for fractional permissions can be
\emph{dramatically} simpler than existing state-of-the-art approaches,
including both industrial languages like Rust, as well as the academic
languages developed by \cite{bierhoff}.  Their type system, much like
Rust, build a complex dataflow analysis into the typing rules to infer
when variables can be shared or not. This allows for very
natural-looking user programs, but creates the impression that using
fractional permissions requires a heavy theoretical and engineering
effort going well beyond that needed for supporting basic linear
types.

Instead, our approach of requiring sharing to be made explicit means
that the existing unification machinery already in place for ordinary
ML-style type inference, can be reused to support fractions. Basically,
we can view sharing a value as dividing a fraction by two, and after
taking logarithms all fractions are Peano numbers, whose equality can
be established with ordinary unification.

@@

However, since our type system shows demonstrates unification is ``all
one needs'' for fractions, it should be possible to \emph{encode}
\lang's approach to fractional permissions in Linear Haskell by adding
a GADT-style natural number index to array types tracking the
fraction, which should enable supporting high-performance BLAS
bindings in Linear Haskell. Actually implementing this is something we
leave for future work, as there remains an issue which we do not see a
good encoding for. Namely, only having support for linear functions
makes it a bit inconvenient to manipulate linear values directly --
programs end up taking on a CPS-like structure. This seems to remain
an advantage of a direct implementation of linear types.

@@

We could also attempt to catch mismatched dimensions at compile time as well.
While this could be done with generative phantom types\cite{abe2015simple},
using dependent types may offer more flexibility in \emph{partitioning}
regions\cite{space_monads} or statically enforcing dimensions related
constraints of the arguments at compile-time.  ATS\cite{cui2005ats} is just
such a language which combines dependent and linear types; although it provides
BLAS bindings, it does not aim to provide aliasing restrictions as demonstrated
in this paper.

>>>>

Using linear types for BLAS routines is a particularly good domain fit
(given the implicit restrictions on aliasing arguments), and as a
result the idea of using substructural types to express array
computations is not a particularly new
one\cite{scholz,henriksen,bernardy2016}.  However, many of these
designs have been focused on building languages to \emph{implement}
the kernel linear algebra functions, and as a result, they tend to add
additional limitations on the language design. Both
Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
higher-order functions to facilitate compilation to GPUs. The work of
\cite{bernardy2016} forbids term-level recursion, in order to ensure
that all higher-order computations can be statically normalized away
and thereby maximize opportunities for array fusion.

@@

Working explicitly with fractional permissions has two main
benefits. First, our type system demonstrates that type systems for
fractional permissions can be dramatically simpler than existing
state-of-the-art approaches, including both industrial languages like
Rust, as well as academic (such as those developed by
\cite{bierhoff}).  Bierhoff \emph{et al}'s type system, much like
Rust's, builds a complex dataflow analysis into the typing rules to
infer when variables can be shared or not. This allows for more
natural-looking user programs, but can create the impression that
using fractional permissions requires a heavy theoretical and
engineering effort going well beyond that needed for supporting basic
linear types.

Instead, our approach, of requiring sharing to be made explicit, lets
us demonstrate that the existing unification machinery already in
place for ordinary ML-style type inference can be reused to support
fractions. Basically, we can view sharing a value as dividing a
fraction by two, and after taking logarithms all fractions are Peano
numbers, whose equality can be established with ordinary unification.

@@

However, since our type system shows demonstrates unification is ``all
one needs'' for fractions, it should be possible to \emph{encode}
\lang's approach to fractional permissions in Linear Haskell by adding
a GADT-style natural number index to array types tracking the
fraction, which should enable supporting high-performance BLAS
bindings in Linear Haskell. Actually implementing this is something we
leave for future work, as there remains one issue which we do not see
a good encoding for. Namely, only having support for linear functions
makes it a bit inconvenient to manipulate linear values directly --
programs end up taking on a CPS-like structure. This seems to remain
an advantage of a direct implementation of linear types over the
Linear Haskell style.

@@

We could also attempt to catch mismatched dimensions at compile time
as well.  While this could be done with generative phantom
types\cite{abe2015simple}, using dependent types may offer more
flexibility in \emph{partitioning} regions\cite{space_monads} or
statically enforcing dimensions related constraints of the arguments
at compile-time.  ATS\cite{cui2005ats} is an example of a language
which combines linear types with a sophisticated proof layer. But
although it provides BLAS bindings, it does not aim to provide
aliasing restrictions as demonstrated in this paper.
